---
title: 'R Shiny app for graphic designers, Create your own pattern '
author: "Serdar Korur"
date: "10/3/2019"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

STRUCTURE:

1. Problem  = Value

Check R Shiny blog
You made explorations of your data

ggplot skills - how to create patterns in ggplot
formulas - 

2. Body

Insights - Conclusion

show: ggplot2, r shiny, web app, embedding, 
-------------

Inspired from a [datacamp project](https://projects.datacamp.com/projects/62) I decided to make a Shiny app to draw patterns and share it online.

With Gpplot2 package you can do visualizations and share insights of your data, but you can also use it for drawing patterns.

Before that, let's look at a bit how ggplot internals work.

## GGPLOT tips: distinction between aesthetics and attributes

Like lego bricks layered on top of each other works in layers.

To use ggplot with its full potential let's make an important distinction between how ggplot understands aesthetics and attributes.

When I talk about Aesthetics those are the terms inside **aes()** in ggplot syntax and attributes are those not.

e.g ggplot(data, aes(x, y, **color=var1**) + geom_point**(size=6)**


Aesthetics typically undertood as how something looks, color, size etc. But in ggplot's world how things look is just an attribute.

{{% alert warning %}}
Aesthetics refer not to refer something looks, **but to which variable is mapped onto it.**
{{% /alert %}}



I will create an imaginary data frame, that we can apply those concepts.

```{r define the dataframe}
points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle
x <- sin(t/2)
y <-cos(t/2)
z <- rep(c(1,2,3,4,5,6,7,8,9,10), times=50)
w <- rep(c(1,2), times=250)
df <- data.frame(t, x, y, z, w)
# Have a look at the data
head(df)
```

Many patterns in Nature can be explained by mathematical terms, Shapes of sunflowers, dandelions or snowflakes etc.

```{r ggplot main spiral}
# Make a scatter plot of points of a spiral
p <- ggplot(df, aes(x*t, y*t))
p + geom_point()
```

# Example use of an aesthetics

By defining col=factor(z) inside aes() I can map colors defining each factor of z onto our points.
```{r Map z on color}
# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, col=factor(z)))
p + geom_point()
```

# Example use of an attribute: How somethings looks. I will make the points bigger by defining size=4. 
```{r main spiral colored}

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, col=factor(z)))
p + geom_point(size = 4)

```

```{r spiral colored t/2}
# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, col=factor(z)))
p + geom_point()
```

# Use shape as an attribute
Here I am changing how something looks like The data point shape change to 24 which defines a filled triangle. But nothing is mapped onto it. It is just an attribute. 

I also added inside aes() **factor(z)**. It is mapped on to how the triangles filled (aesthetics)

```{r spiral colors with triangles}

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, color=factor(w), fill=factor(z)))
p + geom_point(shape=24, size=4)
```
# Using shape as an aesthetics
```{r  aa}
points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle
x <- sin(t)
y <-cos(t)
z <- rep(c(1,2,3,4,5,6,7,8,9,10), times=50)
w <- rep(c(1,2), times=250)
df <- data.frame(t, x, y, z, w)

p <- ggplot(df, aes(x*t, y*t, shape=factor(w), color=factor(z)))
p + geom_point(size=3)
```






Shiny is an R package that makes it easy to build interactive web applications (apps) straight from R. 
\Shiny combines the computational power of R with the interactivity of the modern web.


Building Web Applications in R with Shiny

Shiny App contents:

1. ui <- fluidPage(
2.
3. server
As a Data Scientist easy sharing of your results with others who does not need to know exactly the coding below anlysis R Shiny is a great solution. You can create an app others can explore your results interactively. 

You want to share your results remotely and you want to publish it online.



1. Share results and explore data with others
2. Make a fun app share with friends

How to embed a R shiny app on a static website using blogdown. 

**THE APP WILL BE HERE**

<iframe width="1000" height="1500" scrolling="no" frameborder="no"  src="https://korur.shinyapps.io/designapattern/"> </iframe>

My wife is a graphic designer and I wanted to inspire her by drawing patterns with Math and created an app to draw patterns modifying couple of parameters of the math behind the pattern. 



Find out more about building applications with Shiny here:

>http://shiny.rstudio.com/

>It arrange tons of icons in formations like the petal arrangement pattern in a sunflower.

>You can find more info here. [phyllotaxis](https://en.wikipedia.org/wiki/Phyllotaxis)
>Here is a cool video about the [math](https://www.youtube.com/watch?v=KWoJgHFYWxY) behind.

# The shiny app


>To embed a shiny app on a static site I first need to publish it on shinyapps.io. 

Once you create an account you just follow three very simple instructions on the website to deploy your app. 

Once deployed, you end up with a url, which you can then embed in an Rmarkdown file using one line of html. 

Since my focus today is not on how to write a shiny app, I’m going to go through these steps out of order, starting with how to put an app online.



# Publising a shiny app online


Create an account at: 

https://www.shinyapps.io/

profile > Tokens > Show > Copy 

Final step: Go to RStudio:

install.packages("rsconnect")
library(rsconnect)

paste the copied code which look like below

rsconnect::setAccountInfo(name='yourshinyapps_ID',
                          token='KEY',
                          secret=VALUE)

rsconnect::deployApp('path/to/your/app')

Now to to view it on 

https://your_user_id.shinyapps.io/directory_name/


Deploying (Publishing online)
I hope I’m not insulting anyone’s intelligence here, but it wasn’t immediate obvious to me so I will start by defining “deploy”. When you deploy something, you basically are publishing it online.

Ok, now that that’s out of the way, I can explain how to publish it online. After writing my app, I saved it as an Rmarkdown file (app.Rmd), which I deployed at (https://brettory.shinyapps.io/gender_tech_country/). To get R to recognize an .Rmd file as a shiny app, you need to include one extra line in the header: runtime: shiny

Since the shiny app will be running online and not off of your personal computer, you need to upload the data as well. To do so, just make sure the data is in the same folder (or a subfolder within the main folder) as the folder where you store your app.Rmd file. It would also work to link to data that is alreay online. And be sure to use relative paths (ex: “data/KaggleSurvey.csv”) rathter than absolute paths (ex: “User/Brett/Documents/App/data/KaggleSurvey.csv”). If your shiny app is running locally but doesn’t run when you post it online, it’s probably the paths, but I recommend the Rstudio’s help page to troubleshoot other problems.



# Creating the shiny app

My goal was to create an app with two tabs. The first would let users select which country they are interested in, then show them a bar chart of the number of men and women working in tech in each country. In the second tab, I would have a simple tabulation of number of tech workers per gender and per country.

To create this app, I first needed to prep the data as shown above. Because my data manipulations of the Kaggle survey weren’t very complicated, I just included the above code in my app.Rmd file with echo=FALSE heading each chunk so it would run but wouldn’t produce any output. echo=FALSE is just one of the many useful commands that can be found on this Rmarkdown cheat sheet.

The resulting file can be found on GitHub.

Finally, I created the following shiny app:

Once created and deployed, I simply embed my shiny app in my Rmarkdown file with a little html <iframe src="myurl"> </iframe>

Here you can also control the size of the frame with commands width and height, and additional commands like scrolling and frameborder to make it look a little nicer. The exact command I use to embed the shiny app is:

<iframe width="450" height="400" scrolling="no" frameborder="no"  src="https://brettory.shinyapps.io/gend


It does take a while to load, probably because I do data manipulation within the app file rather than in a separate file, but I will leave that problem for another day!

# Final thoughts.

I have created an app on shiny. Published it on shinyapps.io server. I used the link wrap it in an htlm to the server to embed it in my webpage
If you like my app try to create your patterns. This app can create 420 Billion patterns which you can apply your taste of colors.


Examples_

# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle
x <- sin(t)
y <- cos(t)
df <- data.frame(t, x^2, y^2)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x, y*t))
p + geom_point()


points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle
x <- sin(t)
y <- cos(t)
df <- data.frame(t, x^2, y^2)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x, y*t))
p + geom_point(size =4)


points <- 5000
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle
x <- sin(t)
y <- cos(t)
df <- data.frame(t, x^2, y^2)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x, y+t))
p + geom_point(size =2)

extras
points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle 
x <- sin(t)
y <- cos(t)
z <- rep(c(21,22,23,24,25), times=100)
s <- rep(c(1,2,3,4,5,6,7,8,9,1), times=50)
df <- data.frame(t, x, y, z, s)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t))
p + geom_point(shape=factor(z), fill=factor(z))

```{r yellow dots}

points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle 
x <- sin(t)
y <- cos(t)
z <- rep(c(21,22,23,24,25), times=100)
s <- rep(c(1,2,3,4,5,1,2,3,4,5), times=50)
df <- data.frame(t, x, y, z, s)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, col=factor(z), size=factor(s)))
p + geom_point(shape=21, fill="yellow")
```


```{r ciurcles colors}
points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle 
x <- sin(t)
y <- cos(t)
z <- rep(c(1,2,23,24,25), times=100)
s <- rep(c(1,2,3,4,5, 6,7,8,9,10), times=50)
df <- data.frame(t, x, y, z, s)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, size=factor(s)))
p + geom_point(shape=21, col=factor(s)) + theme_void()




```{r tents}
points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle 
x <- sin(t)
y <- cos(t)
z <- rep(c(1,2,23,24,25), times=100)
s <- rep(c(1,2,3,3,3,2,1,1,2,1), times=50)
df <- data.frame(t, x, y, z, s)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, size=factor(s)))
p + geom_point(shape=24, fill=factor(s), col=factor(z)) + theme_void()
```







```{r colorful}
points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle 
x <- sin(t)
y <- cos(t)
z <- rep(c(1,2,3,4,5,6,7,8,9,10), times=50)
s <- rep(c(1,2,3,4,5,6,7,8,9,10), times=50)
df <- data.frame(t, x, y, z, s)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t),fill=factor(z), col=factor(s))
p + geom_point(size=4, shape=21, fill=factor(z), col=factor(s)) + theme_void()

```



```{r bb}
points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle 
x <- sin(t)
y <- cos(t)
z <- rep(c(1,2,3,4,5,6,7,8,9,10), times=50)
s <- rep(c(1,2,3,4,5,6,7,8,9,10), times=50)
df <- data.frame(t, x, y, z, s)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, size=4,  col=factor(s), fill=factor(z)))
p + geom_point(shape=factor(s)) + theme_void()
```






# What is R Shiny?
```{r  ccc colorful }
points <- 500
# Defining the Golden Angle
angle <- pi*(3-sqrt(5))
t <- (1:points) * angle 
x <- sin(t)
y <- cos(t)
z <- rep(c(1,2,3,4,5,6,7,8,9,10), times=50)
s <- rep(c(1,2,3,4,5,6,7,8,9,10), times=50)
df <- data.frame(t, x, y, z, s)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t, size=t/y,  col=factor(s), fill=factor(z)))
p + geom_point() + theme_void()
```