---
title: Can machine learning change your lifestyle? Predicting Diabetes
description: "predictive modeling algorithms - for learners"
author: ''
date: '2019-10-01'
output:
    html_document:
        highlight: tango
slug: predict-diseases
categories: []
tags:
  - Machine learning
  - Blog
  - Data Science Process
  - Linear Regression
  - Logistic Regression
  - glmnet
  - Randomforest
  - Data Quality control
  - Feature engineering
  - Pima Indians
subtitle: ''
summary: ''
authors: []
lastmod: '2019-10-01T09:09:36+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Would you be taking care of yourself better if your doctor told today that you have **high risk of diabetes?**

Advances in fields, such as omics and internet of things (sensors that collect data), and centralization of healthcare information (e.g. [OMOP common data model](https://www.ohdsi.org/data-standardization/the-common-data-model/)) enable us to access much wider data sources. 

Gaining insights from those we can improve our well being with better healthcare. Some applications of machine learning tools are;

* Diagnosing diseases earlier
* Identifiying drugs with reduced side effects
* Select patient groups responding to an experimental therapy
* Utilize existing therapies better

**Let's look at an example and try to help some doctors in diagnosing their patients.**

# Problem formulation

> Can we build a Machine learning algorithm to predict which patients will develop Diabetes?

The goal is to predict a Binary **Outcome**: ```Diabetes``` vs ```Healthy```
by using 8 medical indicators.

# Overview of the data
  
  
**The Pima Indians of Arizona and Mexico** have contributed to numerous scientific gains. Their involvement has led to significant findings on genetics of both **type 2 diabetes** and **obesity.** 

The medical indicators recorded are;

**Pregnancies:** Number of times pregnant  
**Glucose:** Plasma glucose concentration a 2 hours in an oral glucose tolerance test  
**BloodPressure:** Diastolic blood pressure (mm Hg)  
**SkinThickness:** Triceps skin fold thickness (mm)  
**Insulin:** 2-Hour serum insulin (mu U/ml)  
**BMI:** Body mass index (weight in kg/(height in m)^2)  
**DiabetesPedigreeFunction:** Diabetes pedigree function  
**Age:** Age (years)  
**Outcome:** Class variable (0 or 1)  

# Data acquisition

I downloaded the [Pima Indians Diabetes Dataset](https://www.kaggle.com/uciml/pima-indians-diabetes-database) from Kaggle and imported it from my local directory.

```{r import data and packages, message=FALSE, warning=FALSE}
diabetes <- read.csv("posts_data/diabetes.csv")
library(tidyverse) # ggplot2, dplyr, tidyr, readr, 
                   # purrr, tibble, stringr, forcats
library(reshape2)
library(ggcorrplot)
library(pROC)
library(lattice)
library(caret)
library(waffle)

```


# Data Quality control

Before counting on any algorithm a good starting point is to **check obvious mistakes and abnormalities in your data.**

Here, I would look at **Missing values**, variable ranges (**min, max values**). A very extreme value might be basically a sign of typing error.

## Understand your Data

How big is the data? Classes of variables?
```{r dim}
dim(diabetes)
knitr::kable(sapply(diabetes, class))
```


Next, what catches my attention is **unexpected zero values in Insulin**. Look below.
```{r head}
knitr::kable(head(diabetes))
```

## Missing Values

Summary gives a good overview of the variables. Any missing data will show up here listed as **"NA's"**. 

```{r summary}
summary(diabetes)
```

I will make visualizations of the variables to see how they are distributed.

```{r Check out how variables are distributed}
gg <- melt(diabetes)
ggplot(gg, aes(x=value, fill=variable)) +
  geom_histogram(binwidth=5)+ 
  facet_wrap(~variable) 
```

Peaks at zero of Skin Thickness and Insulin is obvious here. 

In cases where the numbers are small we might remove them. **Let's figure it out with a for loop.** and then visualize on a waffle plot.

This will bring me the number of zero containing rows in variables from 2 to 6.

```{r check number of zeros in each variable}
zero_rows <- list()
for(i in 2:6){
zero_rows[[i]] <- length(which(diabetes[,i] == 0))  
}
rows_with_zero <- unlist(zero_rows)
```

Feed those numbers to a waffle plot.

```{r waffle plot of zero values}
zeros <- c("Glucose" =rows_with_zero[1], "Blood Pressure" = rows_with_zero[2], 
           "Skin Thickness"= rows_with_zero[3], "Insulin" =rows_with_zero[4], 
           "BMI" = rows_with_zero[5])
waffle(zeros, rows=20) + 
  theme(text = element_text(size=15)) + 
  ggtitle("Number of rows with zero")
```

```{r zero rows on a dataframe}
df <- data.frame(rows_with_zero, row.names = names(diabetes[2:6]))
df
```


For instance, **374** Insulin values are zero. Other variables also contain zeros. It is impossible to have Blood Pressure or Glucose levels at 0. It is unlikely that those are simply entry mistakes. It seems missing values are filled with **zeros** in the data collection phase. 

**How to circumvent this?**

### Convert all **zeroes** to **NAs** and then perform **Median Imputation**

Most models require numbers, and can't handle missing data. Throwing out rows is not a good idea since it can lead to biases in your dataset and generate overconfident models.

Median imputation lets you model data with missing values. By replacing them with their medians.

To do this, I need to change zeros to missing values. I will do this for all the predictors which zero is not plausible(columns 2 to 6).

```{r Median Imputation}
for(i in 2:6){
# Convert zeros to NAs
diabetes[, i][diabetes[, i] == 0] <- NA
# Calculate median
median <- median(diabetes[, i], na.rm = TRUE)
diabetes[, i][is.na(diabetes[, i])] <- median
}
```

Check if it really happened.
```{r head data}
knitr::kable(head(diabetes))
```

For instance, I see that zero values in the insulin variable is replaced with median of insulin which is 125. 

**Now, the data is clean and ready for the modeling phase.**

# Modeling the data (build, fit and validate a model)

Before going into any complicated model starting with a simple model is a good idea. It might do surprisingly well and will give us more insights. 

## Logistic Regression Model

We will create two random subsets of our data in 80/20 proportion as **training and test data.** Training data will be used to build our model and test data will be reserved to validate it. 
```{r Create Train/test split}
set.seed(22)
# Create train test split
sample_rows <- sample(nrow(diabetes), nrow(diabetes) * 0.8)
# Create the training dataset
dia_train <- diabetes[sample_rows, ]
# Create the test dataset
dia_test <- diabetes[-sample_rows, ]
```

```{r Build a simple model}
# Build a logistic regression model with the train data
glm_dia <- glm(Outcome ~ .,data = dia_train)
summary(glm_dia)
```

The summary shows us not all the variables play a role in predicting outcome. The signicant correlations was found for Pregnancies, Glucose, BMI and Pedigree function.

The predict function will give us probabilities. To compute our model accuracy we need to **convert them to class predictions by setting a threshold level.**

```{r Predict test data model}
# We will predict the Outcome for the test data
p<-predict(glm_dia, dia_test)
# Choose a threshold 0.5 to calculate the accuracy of our model
p_05 <- ifelse(p > 0.5, 1, 0)
table(p_05, dia_test$Outcome)
```

We will build a confusion matrix to calculate how accurate our model is in this particular random train/test split and at 0.5 threshold level. 

```{r Confusion matrix}
conf_mat <- table(p_05, dia_test$Outcome)
accuracy <- sum(diag(conf_mat))/sum(conf_mat)
accuracy
```

**roc** function pROC package, can plot us a ROC curve which tests accuracy of our model at multiple threshold levels and is a good estimate on how well our model is performing.

```{r Calculate AUC}
# Calculate AUC(Area under the curve)
roc(dia_test$Outcome, p)
```


However, this process is little fragile, presence or absence of a single outlier might vastly change the results you might get from a given random train/test split. 

A better approach than a simple train/test split is using multiple test sets and averaging their accuracies.

Let's test that. I will create 1, 30 or 1000 random test sets, build models and compare their accuracies. 

#### How to apply multiple train/test split

To do this, I will **write a function** where I can choose number of independent train/test splits.  
  
  
It will return me an average value of the accuracy(auc) of the model after chosen number of iteration. The higher the number of random splits the more stable your estimated AUC will be.

Let's see how it will work out for our diabetes patients.

```{r multi_split function, warning=FALSE, message=FALSE}

# I will define my function as follows
multi_split <- function(x){
sample_rows <- list()
dia_train <- list()
dia_test <- list()
glm <- list()
p <-  list()
roc_auc <- list()
for(i in 1:x){
  sample_rows[[i]] <- sample(nrow(diabetes), nrow(diabetes) * 0.8)
  # Create the training dataset
  dia_train[[i]] <- diabetes[sample_rows[[i]], ]
  # Create the test dataset
  dia_test[[i]] <- diabetes[-sample_rows[[i]], ]
  glm[[i]] <- glm(Outcome ~ .,data = dia_train[[i]])
  p[[i]] <- predict(glm[[i]], dia_test[[i]])
  
  # Calculate AUC for all "x" number of random splits
  roc_auc[[i]] <- roc(dia_test[[i]]$Outcome, p[[i]])$auc[1]
  glm_mean <- mean(unlist(roc_auc))
}
print(mean(unlist(roc_auc)))
}
```

Let's calculate the average AUC of our model after different number of random splits.

I will run my **multi_split() function** 3x for 1, 30 and 1000 random train/test splits. I can then compare variances at each level of sampling.

Here are the results from my multi_site function at each randomization.
```{r multi split function test, message=FALSE, warning=FALSE}
auc_1_1 <- multi_split(1)
auc_1_2 <- multi_split(1)
auc_1_3 <- multi_split(1)
auc_30_1 <- multi_split(30)
auc_30_2 <- multi_split(30)
auc_30_3 <- multi_split(30)
auc_1000_1 <- multi_split(1000)
auc_1000_2 <- multi_split(1000)
auc_1000_3 <- multi_split(1000)
```


Let's compare Variance levels at **1, 30 and 1000** random splits

```{r changes in variance, message=FALSE, warning=FALSE}
var(c(auc_1_1, auc_1_2, auc_1_3))
var(c(auc_30_1, auc_30_2, auc_30_3))
var(c(auc_1000_1, auc_1000_2, auc_1000_3))
```

What we see here as we increase the number of iterations to 30 and 1000 the variability
gradually stabilizes around a trustable AUC of ```r round(mean(c(auc_1000_1, auc_1000_2, auc_1000_3)), 3)```. 

Seeing is believing. Let's plot it.

```{r change in variance plot}
# Create a data.frame containing accuracies
random_1X <- c(auc_1_1, auc_1_2, auc_1_3)
random_30X <- c(auc_30_1, auc_30_2, auc_30_3)
random_1000X <- c(auc_1000_1, auc_1000_2, auc_1000_3)

df_r <- data.frame(random_1X, random_30X, random_1000X)
df_r

# Here, I will reformat my data for easy plotting by using gather() function from tidyr
# It takes multiple columns, and gathers them into key-value pairs: it makes “wide” data longer.
df_long <- gather(df_r, sampling, auc)
df_long
df_long$sampling <- factor(df_long$sampling, levels = c("random_1X", "random_30X", "random_1000X"))

# 
model_variation <- ggplot(df_long, aes(y=auc, x=sampling, fill=sampling)) + geom_boxplot() + theme(text = element_text(size=15), axis.title.x=element_blank(), legend.position = "none") + ggtitle("Variation in model performance")
model_variation


```

Great. We have an estimate of our model performance after 1000 random train/test splits. This process is also called **Monte-Carlo Cross validation.** This approach might give you a less variable, but more biased estimate. 

A more common approach to estimate model performance is **k-Fold cross Validation.** Where the samples divided into k-folds and one fold is used as a test set, and the remaining k-1 as the training set. This process is run k times until all folds appear once in the test sample.

## Logistic regression model with k-fold Cross Validation

I will switch here to caret package. With the **Train()** function we can test different types of machine learning algorithms and set the cross validation parameters.


To make the models below comparable I will create **a custom cross validation fold object (d_folds)** that I can apply to multiple models. 

I will repeat the logistic regression model with 5 fold cross validation and then we can compare it to monte carlo cross validation. 

```{r Logistic Regression model function, warning=FALSE, message=FALSE}
# Convert Outcome to a factor with two levels
diabetes$Outcome <- ifelse(diabetes$Outcome == 1, "Yes", "No")
outcome <- diabetes$Outcome
d_folds <- createFolds(outcome, k=5)

# Create a dataframe without the outcome column
diab <- diabetes[,-9]

# MyControl
myControl <- trainControl(
    summaryFunction = twoClassSummary,
    classProbs = TRUE,
    verboseIter = TRUE,
    savePredictions = TRUE,
    index = d_folds
)
# Model_glm
model_glm <- train(x = diab, y = outcome,
                   metric = "ROC",
                   method = "glm",
                   trControl = myControl
)
model_glm
```

Here, My model performance is ```r model_glm$results[1,2]```

## Glmnet model


```{r Glmnet model, warning=FALSE, message=FALSE,  results="hide"}
# Model
model_glmnet <- train(x = diab, y = outcome,
                   metric = "ROC",
                   method = "glmnet", tuneGrid = expand.grid(
                          alpha = 0:1,
                          lambda = seq(0.0001, 1, length = 20)
                      ),
                   trControl = myControl
)
```


```{r print Glmnet model, warning=FALSE, message=FALSE}
model_glmnet
plot(model_glmnet)
```

As we see in the plot, ridge regression (alpha = 0) performed better than the lasso at all lambda values.

Glmnet model performance is ```r max(model_glmnet$results$ROC)```

## Random forest model

One of the big diferences between random forest and linear models is that they require "tuning."

Hyperparameters --> How the model is fit. Selected by hand.

advantages: no need to log transform or normalize,
but they are less interpretable and slower than glmnet.

Random forests **capture threshold effects and variable interactions. both of which occur often in real world data** 

**mtry** is the number of variables used at each split point in individual decison tree that make up the rf. Default is 3, I will use here 8. 

tuneLength = how many different mtry values to be tested. 

```{r Random forest model, warning=FALSE, message=FALSE,  results="hide"}
# Random forest model
model_rf <- train(x = diab, y = outcome,
                   tuneLength = 8,
                   metric = "ROC",
                   method = "ranger",
                   trControl = myControl
)
```


```{r print Random forest model, warning=FALSE, message=FALSE}

model_rf
```

Random forest performance is ```r max(model_rf$results$ROC)```


## Gradient boost model

I will define manualy a grid to test hyperparameter values wider than set in default. 
```{r Gradient boost model, warning=FALSE, message=FALSE, results="hide"}

grid <- expand.grid(interaction.depth = c(1, 2, 3, 4, 5),
                 n.trees = (1:20)*50,  shrinkage = 0.01,
             n.minobsinnode = 10)
model_gbm <- train(x = diab, y = outcome,
                   metric = "ROC",
                   method = "gbm",
                   tuneGrid = grid,
                   trControl = myControl
)
```


```{r print Gradient boost model, warning=FALSE, message=FALSE}
model_gbm
```

Gradient boost model performance is ```r max(model_gbm$results$ROC)```

## Naive Bayes model

```{r Naive Bayes model, warning=FALSE, message=FALSE, results="hide"}
model_nb <- train(x = diab, y = outcome,
                   metric = "ROC",
                   method = "nb",
                   trControl = myControl
)
```


```{r print Naive Bayes model, warning=FALSE, message=FALSE}
model_nb
```

Naive Bayes model performance is ```r max(model_nb$results$ROC)```

```{r Summary}
models <- c("glm", "glmnet", "rf", "gbm", "naive")
glm <- max(model_glm$results$ROC)
glmnet <- max(model_glmnet$results$ROC)
rf <- max(model_rf$results$ROC)
gbm <- max(model_gbm$results$ROC)
naive <- max(model_nb$results$ROC)
AUC <- c(glm, glmnet, rf, gbm, naive)
df <- data.frame(models, AUC)
df<- df[order(df[,2], decreasing=TRUE), ]
knitr::kable(df)
```


Here, we found ```r df[1,1]``` model performed the best, and also there are not big differences between the models.



## Future thoughts

I used different machine learning algorithms to predict Diabetes. Models showed similar performances except the naives bayes which performed worst.  **As we saw, our simple glm model performance was very close to other more advanced algorithms.**   

We can help doctors to predict **Diabetes with accuracy around 83%** by using 8 simple medical parameters.
  
Given current speed in generation and collection of types data by including additional predictors we can build even better models. 

Until next time!

Serdar
 